Red Wine Data Analysis by Sourabh Dev
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(psych)
library(gridExtra)
library(corrplot)
```

```{r echo=FALSE, Load_the_Data}
setwd("/Users/dev/Documents/Udacity/Explore and Summarize Project")
# Load the Data
winedata <- read.csv("wineQualityReds.csv")
```

# Univariate Plots Section

Lets start by looking at the data summary
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
str(winedata)

summary(winedata)
```

From this summary we can see some broad categories like: acidity, sugar, chemical groups, quality, alcohol content.

Lets start by plotting the quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=as.factor(quality)), data = winedata) +
  labs(x='quality') +
  geom_bar()
```
This looks like a normal distribution.

To continue this analysis further, lets look at the: density, alcohol levels and sugar
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density), data = winedata) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol), data = winedata) +
  geom_histogram()
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(winedata$alcohol)

ggplot(aes(x=alcohol), data = winedata) +
  geom_histogram(binwidth = 0.1)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(winedata$alcohol)

ggplot(aes(x=alcohol), data = winedata) +
  scale_x_log10() +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=residual.sugar), data = winedata) +
  geom_histogram(binwidth = 0.1)
```

The density looks like a normal distribution and the alcohol data is a little skewed. We can see a large spike in the alcohol level around 9.5%.

Sugar seems to be skewed drastically, it would make sense to test it on a log scale.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=residual.sugar), data = winedata) +
  scale_x_log10(breaks=c(1, 2, 5, 10)) +
  geom_histogram()
```

Nothing significant can be seen here.

Now, lets look at the acidity
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pH), data = winedata) +
  geom_histogram(binwidth = 0.1)
```

pH seems to follow a normal distribution, with the largest concentration around 3.3.

```{r echo=FALSE, message=FALSE, warning=FALSE}
fixed_acidity_plot <- ggplot(aes(x=fixed.acidity), data = winedata) +
  geom_histogram(fill=I('#099DD9'), binwidth = 0.1)

volatile_acidity_plot <- ggplot(aes(x=volatile.acidity), data = winedata) +
  geom_histogram(fill=I('#F79420'), binwidth = 0.01)

citric_acid_plot <- ggplot(aes(x=citric.acid), data = winedata) +
  geom_histogram(fill=I('#FF0D09'), binwidth = 0.01)

grid.arrange(fixed_acidity_plot, volatile_acidity_plot, citric_acid_plot)
```

Looks like the fixed and volatile acidity seems to skewed. But, no pattern is visible in case of the citric acid levels. So, lets further explore it.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid), data = winedata) +
  scale_x_log10() +
  geom_histogram(fill=I('#FF0D09'))
```
It seems skewed when measured on a log scale.

Finally, lets explore the chemical levels
```{r echo=FALSE, message=FALSE, warning=FALSE}
chloride_plot <- ggplot(aes(x=chlorides), data = winedata) +
  geom_histogram(fill=I('#F79420'), binwidth = 0.001)

sulphate_plot <- ggplot(aes(x=sulphates), data = winedata) +
  geom_histogram(fill=I('#FF0D09'), binwidth = 0.01)

grid.arrange(chloride_plot, sulphate_plot)
```

These plots look like normal distributions if we remove the outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
fsd_plot <- ggplot(aes(x=free.sulfur.dioxide), data = winedata) +
  geom_histogram(fill=I('#F79420'), binwidth = 1)

tsd_plot <- ggplot(aes(x=total.sulfur.dioxide), data = winedata) +
  geom_histogram(fill=I('#FF0D09'), binwidth = 1)

grid.arrange(fsd_plot, tsd_plot)
```
Both distributions are skewed.
# Univariate Analysis

### What is the structure of your dataset?
The are 1599 different wine bottles and the dataset has 13 features ("fixed.acidity","volatile.acidity","citric.acid","residual.sugar","chlorides","free.sulfur.dioxide","total.sulfur.dioxide","density","pH","sulphates","alcohol","quality").

Some interesting observations:
* Majority of the wines are rate a quality of 5 or 6.
* The alcohol levels are skewed with a large spike at 9.5%.
* The median pH values is at 3.31.

### What is/are the main feature(s) of interest in your dataset?
The main feature in this dataset is the quality. 

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The main features of interest are citric.acid, residual.sugar, ph and alcohol. It would be interesting to see how these variables effect the quality.

### Did you create any new variables from existing variables in the dataset?
No.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
Citric acid and Alcohol seem to be a little unusual. Alcohol seems to have a skewed distribution with a sudden did, it's looks almost bimodal. While citric acid is skewed on the log scale along the x axis.

No aditional changes were made.


# Bivariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
analysis_winedata <- winedata[, c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar", "density", "pH", "alcohol", "quality")]
cor(analysis_winedata)
# pairs.panels(winedata)
```
Lets draw a correlation plot to have a better understaing.
```{r echo=FALSE, message=FALSE, warning=FALSE}
corrplot(cor(analysis_winedata), method = "circle")
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
pairs.panels(analysis_winedata)
```

From the above table and plot matrix we see "fixed.acidity", "volatile.acidity" and "pH" has some correlation with "citric.acid". Interestingly, density has some correlation with "fixed.acidity" and "alcohol". Also, "quality" has some correlation with "alcohol".

Lets now look at pH, fixed.acidity and volatile.acidity versus citric.acid.
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y=pH), data = winedata) +
  geom_point()
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y=pH), data = winedata) +
  xlim(0, quantile(analysis_winedata$citric.acid, 0.99)) +
  geom_point(alpha = 0.5, position = 'jitter')

lm(formula = citric.acid ~ pH, data=analysis_winedata)
```

From the scatter plot we can see that the data seems to be slightly negatively correlated.
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y=fixed.acidity), data = winedata) +
  geom_point()
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y=fixed.acidity), data = winedata) +
  xlim(0, quantile(analysis_winedata$citric.acid, 0.99)) +
  geom_point(alpha = 0.5, position = 'jitter')

lm(formula = citric.acid ~ fixed.acidity, data=analysis_winedata)
```

From the scatter plot we can see that the data seems to be slightly positively correlated.
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y=volatile.acidity), data = winedata) +
  geom_point()
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y=volatile.acidity), data = winedata) +
  xlim(0, quantile(analysis_winedata$citric.acid, 0.99)) +
  geom_point(alpha = 0.5, position = 'jitter')

lm(formula = citric.acid ~ volatile.acidity, data=analysis_winedata)
```
This data looks very similar to pH vs citric acid levels. Maybe pH and volatile.acidity have some relationship. Let's try to plot it.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pH, y=volatile.acidity), data = winedata) +
  xlim(min(analysis_winedata$pH), quantile(analysis_winedata$pH, 0.99)) +
  ylim(0, quantile(analysis_winedata$volatile.acidity, 0.99)) +
  geom_point(alpha = 0.5, position = 'jitter')

lm(formula = pH ~ volatile.acidity, data=analysis_winedata)
```
There definitly seems to be some sort of correlation here.

Now, lets look at denisty vs alcohol and density vs fixed.acidity. 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=alcohol), data = winedata) +
  geom_point()
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=alcohol), data = winedata) +
  geom_point(alpha=0.5, position = 'jitter')

lm(formula = density ~ alcohol, data=analysis_winedata)
```
The general trend here seems to be that alcohol levels decrease with density. Which does make sense as alcohol is lighter than water and more alcohol means less water, hence lower density.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=fixed.acidity), data = winedata) +
  geom_point()
```
There is a clearcut linear relationship between fixed acidity and density. The acidity goes up with the density.

Now, lets more to the most interesting plot between alcohol and quality.
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=factor(quality), y=alcohol), data = winedata) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape=4)

describeBy(winedata$alcohol, winedata$quality)
```
There seems to be a positive correlation, except in the case of wines rates 5 in quality.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
Most of the comparisons made with citric acid showed some type of linear realtionship.

The comparision between alcohol and density proved the hypothesis that wines having low alcohol levels have high concentration of water, hence lower higher in density as water is more dense.

Finally, quality and alcohol showed an increasing linear relationship. But, there is a suddent dip in case of wine with quality '5'.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
As mentioned above the dip in quality vs alcohol is very intersting.

### What was the strongest relationship you found?
pH and fixed acidity seem to have the strongest correlation.



# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x = density, y = alcohol, color = factor(quality)), data = analysis_winedata) + 
  geom_point() +
  theme_dark() +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality',
                                          override.aes = list(alpha = 1, size = 2)))
```
In the above plot of Alcohol vs Density vs Quality. We can see that alcohols rated 5 in quality are on the more denser while having low alcohol content.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol, color = factor(quality)), data = subset(analysis_winedata, analysis_winedata$residual.sugar < 4)) + 
  geom_point() +
  theme_dark() +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality',
                                          override.aes = list(alpha = 1, size = 2)))
```
No significant observations can be derived from this plot.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = pH, color = factor(quality)), data = analysis_winedata) + 
  geom_point() +
  theme_dark() +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality',
                                          override.aes = list(alpha = 1, size = 2)))
```
There are no interesting patterns here.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = pH, y = citric.acid, color = factor(quality)), data = analysis_winedata) + 
  geom_point() +
  theme_dark() +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality',
                                          override.aes = list(alpha = 1, size = 2)))
```
Clearly acidity varies negatively with the pH. But, the quality seems to be uniform.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
From the first graph it seems to be that the density has a inverse relationship with quality. Denser the wine, lower it's score.

### Were there any interesting or surprising interactions between features?
No.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
fixed_acidity_plot <- ggplot(aes(x=fixed.acidity), data = winedata) +
  geom_histogram(fill=I('#099DD9'), binwidth = 0.1)

volatile_acidity_plot <- ggplot(aes(x=volatile.acidity), data = winedata) +
  geom_histogram(fill=I('#F79420'), binwidth = 0.01)

citric_acid_plot <- ggplot(aes(x=citric.acid), data = winedata) +
  geom_histogram(fill=I('#FF0D09'), binwidth = 0.01)

grid.arrange(fixed_acidity_plot, volatile_acidity_plot, citric_acid_plot)
```

### Description One
Their seems to be spikes in the citric acid instead of the expected normal distributions.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x=factor(quality), y=alcohol), data = winedata) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape=4)
```

### Description Two
This boxplot shows how quality varies with alcohol level with a dip at '5'.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = residual.sugar, y = alcohol, color = factor(quality)), data = subset(analysis_winedata, analysis_winedata$residual.sugar < 4)) + 
  geom_point() +
  theme_dark() +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality',
                                          override.aes = list(alpha = 1, size = 2)))
```

### Description Three
This plot shows how residual sugar varies with alcohol. Even though the alcohol levels vary widely with sugar, there is a clear preference for wines with lower amount of residual sugar.
------

# Reflection
The take aways from this analysis are that wines with high quality tend to have higher alcohol content and low residual sugar. Another interesting finding was that citric acidity decreases with pH levels. So, wines with lower acidty have higher citric acid content.

In conclusion, if you are looking for a good bottle of wine. It will most like have very little sweetness to it, but will be strong.